Compositional Quantitative Resource Analysis of C Programs

### 2014-06-23 paper/main.tex
new analysis rules
first try at a call rule

\begin{abstract}
  The goal of quantitative resource analysis is to provide developers
  with quantitative information about the runtime behavior of software
  at development time.  Recent years have seen tremendous progress in
  automatically deriving worst-case resource bounds.  Yet, many
  challenges in specifying, interactively deriving, formally
  certifying, and composing resource bounds remain unsolved.

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  First, the
  runtime resource cost of programs is formalized with a parametric
  cost semantics for CompCert Clight.  Second, a quantitative Hoare
  logic that can be used for interactive bound development is
  implement and proved sound with respect to the cost semantics in the
  Coq Proof Assistant.  Third, an automatic amortized resource
  analysis for C programs that can compute derivations in the
  quantitative Hoare logic is developed and implemented.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.

  The practicability of the analysis framework is demonstrated with an
  experimental evaluation.  The logic is used to derive XXX A comparison
  with other automatic tools shows that the constant factors in the
  bounds that are derived by the automatic amortized analysis are
  often more precise. XXX

@@

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other tool is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
computes the bound $x + y + 10$, PUBS~\cite{AlbertAGPZ12} computes
the tight bound $0.1(y-x)$, Rank~\cite{AliasDFG10} computes the bound XXX, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.

@@

To understand the capabilities of the automatic analysis, it is
instructive to study more involved examples.  \pref{fig:ex_list}
contains Examples \emph{speed\_1} and \emph{speed\_2} that are taken
from previous work on bound analysis~\cite{GulwaniMC09}.  The SPEED
tool~\cite{GulwaniMC09} derives the bounds as our analysis but
requires inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

@@

Example \emph{t27} shows how amortization can be used to handle tricky
nested loops.  The outer loop we increment the variable $n$ till $n =
0$.  In each of the $|[n,0]|$ iterations, we increment the variable
$y$ by $1000$.  We then non-deterministically execute an inner loop
that decrements $y$ by $100$ until $y<100$.  The analysis can derive
that only the first execution of the inner loop depends on the initial
value of $y$.  We again derive tight constant factors.

As mentioned, the analysis naturally handles advanced control flow
like \code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system can compute a tight bound.  On this example, the
analysis amounts to computing the meeting point of two trains that
approach each other with different speeds.
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

We only show a very small selections of the programs that we can
handle automatically here.  In \pref{app:cat} you find a list of more
than $30$ classes of challenging programs that we can automatically
analyze.  We hope that this gives you an impression of the
capabilities of the analysis but we also point out that this list is
far from being exhaustive.

>>>>

\begin{abstract}

@@

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+1}.  To the best of our knowledge, no other tool is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
computes the bound $x + y + 10$, PUBS~\cite{AlbertAGPZ12} computes
the tight bound $0.1(y-x)$, Rank~\cite{AliasDFG10} computes the bound XXX, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.

@@

To understand the capabilities of the automatic analysis, it is instructive to
look at some more involved examples.  Examples \emph{speed\_1} and
\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.  They
are both handled by the SPEED tool~\cite{GulwaniMC09} but require
inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

@@

Example \emph{t27} shows how amortization can be used to handle tricky
nested loops.  The outer loop is iterated $|[n,0]|$ times.  In the
outer loop, we increment the variable $n$ till $n = 0$.  In each of
the $|[n,0]|$ iterations, we increment the variable $y$ by $1000$.  We
then non-deterministically execute an inner loop that decrements $y$
by $100$ until $y<100$.  The analysis can derive that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive a tight bound.

As mentioned, the analysis naturally handles advanced control flow
like \code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system can compute a tight bound.  On this example, it
amounts to computing the meeting point of two trains that approach
each other with different speeds.
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

Please note that we can only show a very small selections of the
programs that we can handle automatically here.  In \pref{app:cat} you
find a list of more than $30$ classes of challenging programs that we
can automatically analyze.  We hope that this gives you an impression
of the capabilities of the analysis but we also point out that this
list is far from being exhaustive.
